Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-11-01 doc/paper-lcr.tex
conclusions sharpened up

% \voffset=-0.8in

@@

The probability density function (PDF) for the external convergence
given our available data $\data$, $\Pr(\xkappa|\data)$,  is the  key
quantity we seek, since by marginalising over this distribution we can
correctly propagate the uncertainties due to line-of-sight perturbations
(as \citeauthor{FadelyEtal2009} did).  
%
\citet{GunnarssonEtal2006} investigated a related PDF,  for the
magnification, using a galaxy halo model. Applying empirical galaxy
scaling relations both to simulate mock catalogs and then to reconstruct
the line of sight mass distribution. While they did not explore highly
overdense lines of sight, or the impact of groups and clusters, they
found that under their simplifying assumptions, the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gamma|\data)$ from their survey data, running similar Monte Carlo
reconstructions of the mass in nine lens fields based on
\citeauthor{WilliamsEtal2006} and \citeauthor{MomchevaEtal2006}'s
photometric and spectroscopic measurements of galaxies close to the line
of sight. They then  compared the resulting predicted shear
distributions with the external shear demanded by the strong lens model.
Their reconstructions showed most of the shear to have been generated by
bright galaxies within 2~arcminutes of the lens, and that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear. They found significant
discrepancies between the lens model shear and the Monte Carlo
predictions, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies.
%
\citet{KarpenkaEtal2012} used a halo model to predict the magnification due to
galaxies along the line of sight to type Ia supernovae, and inferred the
parameters of the scaling relations between light and mass in these galaxies
from hundreds of supernova lightcone simultaneously. Following
\citet{JonssonEtal2010}, they subtract off the mean convergence to ensure that
the universal mass budget is balanced, and compare models where the halos are
all have either truncated singular isothermal profiles, or 
NFW profiles. Both groups detect lensing effects in the supernova data.

@@

We first consider a purley random selection of lines of sight. With the purely
photometric reconstruction, there is a small bias towards underestimating
$\kappa$: combining 6 targets the half-width and bias of $\mathcal{P}_{6}$
both equal 0.007. With targeted spectroscopic follow up this improves
slightly, for the combination of 8 sightlines the half-width and expectation
value  $\mathcal{P}_{8}$ equal 0.005.  If these lines of sight hosted time
delay lenses, the latter number would correspond to a systematic error of
0.5\% in time delay distance. Since strong lens modelling uncertainties are
typically at the 3-4\% level, a sample of 8 lenses would still be limited by
modelling uncertainty. However, a sample containing 100 or more lenses or more
would be limited by the external convergence reconstruction instead.

\subsubsection{Selecting only the lines of sight with tight $\pr(\kappax)$}
\label{sec:bias:tightPDF}

@@

We find that selecting the most tightly constrained lines of sight {\it
decreases} the bias -- our reconstruction is most \succesful for the lowest
$\kappa$ lines of sight. With photometry alone, the bias in this sample is
0.0028. With targeted spectroscopic follow-up of the selected lines of sight
there is negligible improvement, since the best constrained lines of sight are
the most empty, and it is the high-$\kappa$ lines of sight that benefit the
most from spectroscopy. Photometry of the field is sufficient to adequately
select the best lines of sight for cosmology -- and doing so greatly increases
both the precision and accuracy of the cosmological constraints per system.

@@

By selecting only lines of sight with an external shear of $\gamma>0.03$ we
can test the impact of shear selection. We find that in this sample, $\kappa$
is systematically underestimated at the $\sim$0.01 level. Part of this effect
is mitigated if only the lines of sight with both $\gamma>0.03$ and sharp
$\pr(\kappax)$ are used but biases at the $\sim$0.005 level remain. Since our
model does not include shear constraints it is not surprising that selecting
based on shear can induce a bias. A more sophisticated model that includes the
shear recovered from the lens modelling may be less susceptable to this bias.
The toy model of this section ($\gamma>0.03$) is likely to be a much stronger
selection than reality, but it is worth noting that the reconstruction
\proceedure will be biased if lens selection functions are extreme and
unaccounted for.

@@

We have shown that reconstructing the matter due to halos along any line
of sight can improve the constraints on the external convergence along
that line of sight. The total convergence along a line of sight is
strongly correlated with the reconstructed $\kappah$. However since our
model ignores voids and assumes all halos follow a spherical
truncated-NFW profile our halo model does not include all of the relevant
physics; hence, the width of our resulting $\pr(\kappa|\data)$ is still
typically $\sim$0.01 for any given lightcone, even with a perfect
knowledge of every halo's virial mass and redshift. To make further
progress a more advanced treatment of both voids and halos will be
necessary. \citet{KarpenkaEtal2012} find that their supernova data prefer, in
the context of their simple halo model, a truncated isothermal profile for the
galaxy mass distributions. Weak galaxy-galaxy and cluster-galaxy 
lensing studies
\citep[\eg][]{GavazziEtal2006,JohnsonEtal2007} also suggest an approximately
isothermal profile for the mass-shear correlation function; however, at large
radii these profiles are dominated by the effects of neighbouring halos, which
should be taken into account by our model. Further iterations on the analysis
presented here should include the stellar mass distributions that cause the
isothermal density profiles in the centres of galaxies, although we do not
expect this to have a big impact on the model. Likewise,  halo
ellipticity may have some small effect on the predicted convergence.

@@

The reduction of $\pr(\kappah|\data)$ to just its median value, before using
it to look up the appropriate $\pr(\kappa)$ from the simulated sightline
ensemble, is reminiscent of the procedure followed by \citet{SuyuEtal2010},
and Greene et al (in prep.) The reconstruction method can be seen as the
limiting case of the weightings explored by Greene et al: rather than using
weighted number counts in an aperture, we treat each galaxy individually, and
predict  $\kappa_{\rm h,obs}^{\rm med}$ for use in their place. One
significant disadvantage of these types of method is that they impose a
bottleneck in the flow of information. Once you have compressed to one number,
it is difficult to incorporate more information later; you have to recompute
all the calibrations. Using the full PDF for $\kappah$ is the more correct
thing to do: at present, however, it suffers from significant bias in the
photometric data case. \Fref{fig:no-median-tricks} shows how $\kappa$ is
overestimated, as a result of the high scatter in halo mass for objects with 
high stellar mass. Future work should increasing the precision of the halo
mass estimates.

More generally, the \MS as a calibration tool. If the real Universe is not
like the \MS, then the calibration will be incorrect; repeating the study using
a different simulation to make the test catalogs would allow us to quantify
the sensitivity of our results to the simulation used. Similarly the use of a
different semi-analytic model to paint galaxies onto halos, and an alternative
stellar mass--halo mass relation would both enable similar exploration of
these systematic errors.

More generally, the method presented here is tied to the simulations at
several points: in particular, the prior PDF for $\kappah$, and the
conditional distribution $\pr(\kappa|\kappah)$ used to convert halo model
convergence to total convergence. The former tie could potentially be avoided
if the halo mass estimation gave rise to a reasonable $\pr(\kappah|\data)$ on
its own. Possibilities for improving the precision of the halo mass estimation
are discussed above. The latter tie is more difficult, since breaking it means
modelling the unseen mass structures and voids from first principles. It
should be possible to infer hierarchically  the parameters of a flexible model
for these things from the data itself, as demonstrated by
\citet{KarpenkaEtal2012} -- how much this dilutes the cosmographic precision
is to be investigated.  Including more information in this analysis is key:
weak lensing shears will be available over the whole Euclid and LSST survey
areas: fitting these with the halo models described here in order to constrain
the hyper-parameters of the model is an interesting
possibility.

@@

\item  For the most overdense lines of sight, the reconstruction produces a
very broad PDF; conversely, we find that the lines of sight with the sharpest
inferred $\pr(\kappa|\data)$ are typically under-dense. Since the
reconstructions described here can be performed before follow-up time is
invested, it will be possible to  select targets by the precision of their
$\kappa$ estimates. This procedure results in a sample with low bias as well
as high precision.

@@

\item With only a photometric reconstruction of all time delay lens fields,
and following up only the systems with the most precise $\kappa$ estimates,
the width of the dominant statistical uncertainty in time-delay cosmography
can be halved, without inducing biases at the one percent level.

@@

External convergence is not the only uncertainty in time delay cosmology, but
for systems like B1608$+$656 it is the dominant one. For most lines of sight
it is likely that lens modelling induces uncertainties at the $\sim$3 percent
level, which far exceeds the typical $\kappa$ uncertainties after applying our
reconstruction \proceedure. By using our method to reconstruct lines of sight
and select which lense are followed-up, the uncertainties due to line of sight
convergence will become a sub-dominant  source of error.

>>>>

\voffset=-0.6in

@@

The probability density function (PDF) for the external convergence
given our available data $\data$, $\Pr(\xkappa|\data)$,  is the  key
quantity we seek, since by marginalising over this distribution we can
correctly propagate the uncertainties due to line-of-sight perturbations
(as \citeauthor{FadelyEtal2009} did).  
%
\citet{GunnarssonEtal2006} investigated a related PDF,  for the
magnification, using a galaxy halo model. Applying empirical galaxy
scaling relations both to simulate mock catalogs and then to reconstruct
the line of sight mass distribution. While they did not explore highly
overdense lines of sight, or the impact of groups and clusters, they
found that under their simplifying assumptions, the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gamma|\data)$ from their survey data, running similar Monte Carlo
reconstructions of the mass in nine lens fields based on
\citeauthor{WilliamsEtal2006} and \citeauthor{MomchevaEtal2006}'s
photometric and spectroscopic measurements of galaxies close to the line
of sight. They then  compared the resulting predicted shear
distributions with the external shear demanded by the strong lens model.
Their reconstructions showed most of the shear to have been generated by
bright galaxies within 2~arcminutes of the lens, and that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear. They found significant
discrepancies between the lens model shear and the Monte Carlo
predictions, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies.
%
\citet{KarpenkaEtal2012} used a halo model to predict deterministically 
the magnification due to
galaxies along the line of sight to type Ia supernovae, and inferred the
parameters of scaling relations between light and mass in these galaxies
from hundreds of supernova lightcone simultaneously. Following
\citet{JonssonEtal2010}, they subtract off the mean convergence to ensure that
the universal mass budget is balanced, and compare models where the halos are
all have either truncated singular isothermal profiles, or 
NFW profiles. Both groups were able to 
detect lensing effects in the supernova data this way.

@@

We first consider a purely random selection of lines of sight. With the purely
photometric reconstruction, there is a small bias towards underestimating
$\kappa$: combining 6 targets the half-width and bias of $\mathcal{P}_{6}$
both equal 0.007. With targeted spectroscopic follow up this improves
slightly, for the combination of 8 sightlines the half-width and expectation
value  $\mathcal{P}_{8}$ equal 0.005.  If these lines of sight hosted time
delay lenses, the latter number would correspond to a systematic error of
0.5\% in time delay distance. Since the statistical distance uncertainties due
to lens modelling and time delay estimation are likely to be at the 3-4\%
level \citep{SuyuEtal2010},  only a sample containing 50 or more lenses would
be limited by the residual bias  following external convergence
reconstruction.

\subsubsection{Selecting only the lines of sight with tight $\pr(\kappa|\data)$}
\label{sec:bias:tightPDF}

@@

We find that selecting the most tightly constrained lines of sight {\it
decreases} the bias -- our reconstruction is most \succesful for the lowest
$\kappa$ lines of sight. With photometry alone, the bias in this sample is
0.0028. With targeted spectroscopic follow-up of the selected lines of sight
there is negligible improvement, since the best constrained lines of sight are
the most empty, and it is the high-$\kappa$ lines of sight that benefit the
most from spectroscopy. Photometry of the field is sufficient to adequately
select the best lines of sight for cosmology -- and doing so greatly increases
both the precision and accuracy of the cosmological constraints per system.
In the context of time delay distance accuracy given in the previous section,
the sample size would have to be 160 or so before this residual systematic
error was equal to the statistical uncertainty.

@@

By selecting only lines of sight with an external shear of $\gamma>0.03$ we
can test the impact of shear selection. We find that in this sample, $\kappa$
is systematically underestimated at the $\sim$0.01 level (corresponding to a
1\% systematic error in distance, significant in a sample of 12 lenses). 
Part of this effect is mitigated if only
the lines of sight with both $\gamma>0.03$ and sharp $\pr(\kappa|\data)$ are
used, but biases at the $\sim$0.005 level remain. Since our model does not
include shear constraints it is not surprising that selecting based on shear
can induce a bias. A more sophisticated model that includes the shear
recovered from the lens modelling may be less susceptible to this bias. The
cartoon selection model used here ($\gamma>0.03$) is likely to be a much
stronger selection than reality, but nevertheless it is worth noting that the
reconstruction \proceedure will be biased if lens selection functions are
extreme and unaccounted for.

@@

We have shown that reconstructing the matter due to halos along any line of
sight can improve the constraints on the external convergence along that line
of sight, with, depending on sample selection,  the width of
$\pr(\kappa|\data)$ being up to a factor of two smaller than the global
$\pr(\kappa)$, and  we have quantified the residual bias in the $\kappa$
estimates under our calibration assumptions. We now revisit these assumptions.

The total convergence along a line of sight is strongly correlated with the
reconstructed $\kappah$. However, since our model ignores voids and assumes
all halos follow a spherical truncated-NFW profile our halo model does not
include all of the relevant physics; hence, the width of our resulting
$\pr(\kappa|\data)$ is still typically $\sim$0.01 for any given lightcone,
even with a perfect knowledge of every halo's virial mass and redshift. To
make further progress a more advanced treatment of both voids and halos will
be necessary. \citet{KarpenkaEtal2012} find that their supernova data prefer,
in the context of their simple halo model, a truncated isothermal profile for
the galaxy mass distributions. Weak galaxy-galaxy and cluster-galaxy  lensing
studies \citep[\eg][]{GavazziEtal2006,JohnstonEtal2007} also suggest an
approximately isothermal profile for the mass-shear correlation function;
however, at large radii these profiles are dominated by the effects of
neighbouring halos, which should be taken into account by our model. Further
iterations on the analysis presented here should include the stellar mass
distributions that cause the isothermal density profiles in the centres of
galaxies, although we do not expect this to have a big impact on the model.
Likewise,  halo ellipticity may have some small effect on the predicted
convergence.

@@

The reduction of $\pr(\kappah|\data)$ to just its median value, before using
it to look up the appropriate $\pr(\kappa)$ from the simulated sightline
ensemble, is reminiscent of the procedure followed by \citet{SuyuEtal2010},
and Greene et al (in prep.) The reconstruction method can be seen as the
limiting case of the weightings explored by Greene et al: rather than using
weighted number counts in an aperture, we treat each galaxy individually, and
predict  $\kappa_{\rm h,obs}^{\rm med}$ for use in their place. One
significant disadvantage of these types of method is that they impose a
bottleneck in the flow of information. Once you have compressed to one number,
it is difficult to incorporate more information later; you have to recompute
all the calibrations. Using the full PDF for $\kappah$ is the more correct
thing to do: at present, however, it suffers from significant bias in the
photometric data case. \Fref{fig:no-median-tricks} shows how $\kappa$ is
overestimated, as a result of the high scatter in halo mass for objects with 
high stellar mass. As already noted, 
future work should increasing the precision of the halo
mass estimates.

In this work, we have used 
the \MS as a calibration tool. If the real Universe is not
like the \MS, then this calibration will be incorrect. Repeating this 
study using
a different simulation to make the test catalogs would allow us to quantify
the sensitivity of our results to the simulation used. Similarly the use of a
different semi-analytic model to paint galaxies onto halos, and an alternative
stellar mass--halo mass relation would both enable similar exploration of
these systematic errors.

In particular, the method presented here is tied to the simulations at two
points: the prior PDF for $\kappah$, and the conditional distribution
$\pr(\kappa|\kappah)$ used to convert halo model convergence to total
convergence. The first tie could potentially be avoided if the halo mass
estimation gave rise to a reasonable $\pr(\kappah|\data)$ on its own.
Possibilities for improving the precision of the halo mass estimation are
discussed above. The second tie is more difficult, since breaking it means
modelling the unseen mass structures and voids from first principles. It
should be possible to infer hierarchically  the parameters of a flexible model
for these things from the data itself, as demonstrated by
\citet{KarpenkaEtal2012} (albeit under the assumption of scatter-free scaling
relations) -- how this affects the cosmographic precision is to be
investigated.  In any case, 
including more information in this analysis will be important.
When working with wide field imaging surveys,  weak lensing shears will be
available around every line of sight:  fitting these with the halo models
described here in order to constrain the hyper-parameters of the model is an
interesting possibility.

@@

\item  For the most overdense lines of sight, the reconstruction produces a
very broad PDF; conversely, we find that the lines of sight with the sharpest
inferred $\pr(\kappa|\data)$ are typically under-dense. Since the
reconstructions described here can be performed before follow-up time is
invested, it will be possible to  select targets by the precision of their
$\kappa$ estimates.

@@

\item Selecting the highest precision 33\% of the reconstructed lines of sight
results in a sample with low bias, 0.3\% in terms of the time delay distance,
and a statistical uncertainty due to line of sight mass structure that is half
that of the parent sample. The residual systematic error would be equal to the
overall statistical uncertainty in a sample of around 160 lenses.

\item Conversely, selecting lines of sight with high shear ($|\gamma| >
0.03$ (as might be done inadvertently by focusing on four-image or wide
separation lenses) results in a sample with bias of 1\% in time delay
distance. Such a systematic error would become significant in a sample of 12
lenses.

% \item With only a photometric reconstruction of all time delay lens fields,
% and following up only the systems with the most precise $\kappa$ estimates,
% the width of the dominant statistical uncertainty in time-delay cosmography
% can be halved, without inducing biases at the one percent level.

@@

% PJM: Im not sure this is the right note to end on.

% External convergence is not the only uncertainty in time delay cosmology, but
% for systems like B1608$+$656 it is the dominant one. For most lines of sight
% lens modelling induces uncertainties at the $\sim$3 percent
% level, which far exceeds the typical $\kappa$ uncertainties after applying our
% reconstruction \proceedure. By using reconstruction methods such as that
% explored here to select the most appropriate 
% lenses for follow-up, the uncertainties due to line of sight
% convergence will become a sub-dominant source of error.

While we have tested our reconstruction method against realistic simulations
of line of sight mass distributions in the Universe, our method is also 
calibrated to these simulations. Testing this assumption in the short term,
and replacing it with an empirical calibration (or hierarchical inference) in
the longer term, are worthy goals for future work. The halo model framework is
flexible enough to enable many such improvements, including, quite naturally,
the incorporation of more information from observations.
